Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2018-10-18 WholePlotTrajectories.tex
Jecol review

\Authors{
Ariane MIRABEL\textsuperscript{1*}\\ Bruno Hérault\textsuperscript{2}\\ Eric Marcon\textsuperscript{1}
} % Authors
\affiliation{
\textsuperscript{1}UMR EcoFoG, AgroParistech, CNRS, Cirad, INRA, Université des Antilles,
Université de Guyane.\\ \hspace{1em} Campus Agronomique, 97310 Kourou, France.\\\textsuperscript{2}INPHB, Institut National Polytechnique Félix Houphoüet-Boigny\\ \hspace{1em} Yamoussoukro, Ivory Coast.
}
\affiliation{*\textbf{Corresponding author}: ariane.mirabel@ecofog.gf, https://github.com/ArianeMirabel} % Corresponding author

\Keywords{Community Ecology, Determinants of Plant Community Diversity and Structure, Disturbance Trajectories, Intermediate Disturbance Hypothesis, Long-term Resilience, Neotropical Forests, Taxonomic and Functional Biodiversity} % Keywords - if you don't want any simply remove all the text between the curly brackets
\newcommand{\keywordname}{Keywords} % Defines the keywords heading name

@@

\Abstract{
Understanding the ecological rules underlying the maintenance of
tropical forests biodiversity, structure, functioning and dynamics is
urgent to anticipate their fate in the global change context. The huge
diversity of tropical forests is often assumed to be regularly reshaped
by natural disturbance yielding a diversity peak at intermediate
intensity. This intermediate disturbance hypothesis (IDH), though,
remains debated and the controversy questions the extent of communities'
resilience regarding their taxonomic and functional facets. To
disentangle the ecological processes driving community response to
disturbance, we analyzed the tree community trajectories over 30 years
following a disturbance gradient in a Neotropical forest. Specifically,
we examined community functional and taxonomic trajectories with regards
to diversity, composition and redundancy. Functional trajectories were
drawn based on 7 leaf, stem and life-history traits. We highlighted the
cyclic recovery of community taxonomic and functional composition. While
pre-disturbance taxonomic differences were maintained over time,
functional composition trajectories were quite similar among
communities. The IDH did predict communities taxonomic diversity
response while functional diversity was enhanced whatever the
disturbance intensity. Although consistent, the recovery of community
composition, diversity and redundancy remained unachieved after 30
years. This acknowledged the need of decades-long cycles with no
disturbance to ensure a complete recovery, and questioned tropical
forest community resilience after repeated disturbances.
}

@@

The large areas covered with tropical forests worldwide hold crucial
environmental, economic and social values. They provide wood and
multiple non-timber forest products, shelter a diversified fauna,
regulate the local and regional climates, the carbon, water and nutrient
cycles, and ensure cultural and human well-being. The growing demand in
forests products together with current global changes increases the
pressure on remaining natural forests \citep{Morales-Hidalgo2015} and
threatens the maintenance and dynamics in space and time of communities
structure, composition and functioning \citep{Anderson-Teixeira2013}.

In tropical forests, ecological communities are regularly re-shaped by
natural disturbance events changing both the abiotic environment,
through the fluxes of light, heat and water \citep{Goulamoussene2017},
and the biotic interactions such as competition among species
\citep{Chesson2000}. One of the cornerstone of tropical forest ecology
is to understand the processes and drivers of ecosystems response to
disturbance \citep{Chazdon2003a}. For now, this has been largely studied
through forest structural parameters such as aboveground biomass, tree
height or stem density \citep{Piponiot2016, Rutishauser2016} that are
rapid and convenient to measure. These structural parameters have been
successfully modeled, giving important insights into the recovery of
ecosystem processes and services \citep{Herault2018}. However the
response of forests diversity and composition remains unclear, albeit it
determines the productivity, stability and functioning of ecosystems
\citep{Tilman2014, Liang2016}. In the short-term, moderate disturbance
may lead to positive impacts on communities diversity, an idea
formalized by the intermediate disturbance hypothesis (IDH) stating a
maximized species diversity when disturbance intensity is not too high
\citep{Molino2001, Kariuki2006a}.

Validations of the IDH though remain scarce in the long-term and mainly
rely on the analysis of taxonomic richness \citep{Molino2001}. Taxonomic
richness alone, however, gives limited or misleading information on
forests recovery and functioning \citep{Chaudhary2016}. More
ecological-meaningful analysis would couple richness with (i) evenness,
that would reveal the changes in the species abundance distribution and
thus the underlying ecological processes, and (ii) composition that is
crucial for conservation issues \citep{Lavorel2002, Bellwood2006}.
Furthermore, a functional approach accounting for species biological
attributes would directly link communities diversity, composition and
redundancy to ecosystem functioning and to its environmental constraints
\citep{Violle2007b, Baraloto2012a}. In that respect, the functional
trait-based approach that focus on major traits related to species
ecology and mediate species performance in a given environment was
successfully adopted \citep{Diaz2005}. For instance, the functional
approach revealed in tropical rainforests the deterministic processes
entailing, after disturbance, a functional shift from a dominance of
``conservative'' slow-growing species dealing with scarce resources to
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{Reich2014, Herault2011}. This shift is
translated into the trajectories of key functional traits related to
resource acquisition (leaf and stem traits) and life-history traits
(seed mass, maximum size)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}. Eventually
a complete overview of community response to disturbance would encompass
the changes in functional redundancy, that quantifies the amount of
shared trait values among species \citep{Carmona2016}. The high
functional redundancy of hyper-diverse tropical forests
\citep{Bellwood2006} mitigates the impacts of species removal on
ecosystem functioning and determines the resilience of communities after
disturbance \citep{Elmqvist2003, Diaz2005}.

In this study, we monitored over 30 years the response of 75 ha of
Neotropical forest plots set up on a gradient of disturbance intensity,
from 10 to 60\% of ecosystem above-ground biomass (AGB) loss. We made
use of a large functional traits database encompassing major leaf, stem
and life-history traits in order to draw the taxonomic and functional
trajectories in terms of richness, evenness, composition and redundancy.
Specifically, we (i) elucidated community taxonomic and functional
recovery and the underlying ecological processes, (ii) clarified the
validity of the IDH in the long term for tropical forest and its
translation into different trajectories in time, and (iii) questioned
community recovery time.

@@

Paracou station in French Guiana (5\textdegree 18'N and
52\textdegree 53'W) is located in a lowland tropical rain forest in a
tropical wet climate with mean annual temperature of 26\textdegree C,
mean annual precipitation averaging 2980 mm.y\textsuperscript{-1} (30-y
period) and a 3-month dry season (\textless{} 100
mm.month\textsuperscript{-1}) from mid-August to mid-November, and a
one-month dry season in March \citep{Wagner2011}. Elevation ranges
between 5 and 50 m and soils correspond to thin acrisols over a layer of
transformed saprolite with low permeability generating lateral drainage
during heavy rains.

The experiment is a network of twelve 6.25ha plots that underwent a
disturbance gradient of three logging, thinning and fuelwood cutting
treatments (Table \ref{tab:Tab1}) according to a randomized plot design
with three replicate blocks of four plots. The disturbance corresponds
to averages of 10 trees removed per hectare with a diameter at 1.3 m
height (DBH) above 50 cm for treatment 1 (T1), 32 trees/ha above 40 cm
DBH for treatment 2 (T2) and 40 trees/ha above 40 cm DBH for treatment 3
(T3). Treatments T2 and T3 besides included the thinning of trees by
poison girdling \citep{Schmitt1990}. The disturbance intensity was
measured as the percentage of aboveground biomass (\%AGB) lost between
the first inventory in 1984 and five years after disturbance
\citep{Piponiot2016} estimated with the BIOMASS R package
\citep{Biomass2018}.

@@

The study site corresponds to a tropical rainforest typical of the
Guiana Shield with a dominance of Fabaceae, Chrysobalanaceae,
Lecythidaceae and Sapotaceae. In the twelve experimental plots, all
trees above 10 cm DBH have been mapped and measured annually since 1984.
Trees are first identified with a vernacular name assigned by the forest
worker team, and afterward with a scientific name assigned by botanists
during regular botanical campaigns. In 1984, specific vernacular names
were given to 62 commercial or common species whereas more infrequent
ones were identified under general identifiers only distinguishing trees
and palm trees. From 2003, botanical campaigns have been conducted every
5 to 6 years to identify all trees at the species level but
identification levels still varied among plots and campaigns.

This variability of protocols in time raised methodological issues as
vernacular names usually correspond to different botanical species. It
resulted in significant taxonomic uncertainties that had to be
propagated to composition and diversity metrics. The uncertainty
propagation was done through a Bayesian framework reconstituting
complete inventories at genus level from real incomplete ones on the
basis of vernacular/botanical names association. Vernacular names were
replaced through multinomial trials based on the association probability
\(\big[\alpha_1, \alpha_2,..., \alpha_V\big]\) observed across all
inventories between each vernacular name \emph{v} and all species
\(\big[s_1, s_2,..., s_N\big]\):

@@

Six functional traits representing leaf economics (leaves thickness,
toughness, total chlorophyll content and specific leaf area) and stem
economics (wood specific gravity and bark thickness), and life-history
traits (maximum specific height and seed mass) came from the BRIDGE
project \footnote{http://www.ecofog.gf/Bridge/}. Trait values were
assessed from a selection of individuals located in nine permanent plots
in French Guiana, including two in Paracou, and comprised 294 species
pertaining to 157 genera. Missing trait values (10\%) were filled using
multivariate imputation by chained equation \citep{Mice2011}.
Imputations were restricted within genus or family when samples were too
scarce, in order to account for the phylogenetic signal. Whenever a
species was not in the dataset, it was attributed a set of trait values
randomly sampled among species of the next higher taxonomic level (same
genus or family). As seed mass information was classified into classes,
no data filling process was applied and analyses were restricted to the
414 botanical species recorded.

@@

To counter taxonomic uncertainties due to the variability of botanical
identification levels (in space) and protocols (in time), the taxonomic
composition and diversity analysis were conducted at the genus level.
Taxonomic and functional trajectories of community composition were
followed in a two-dimensional NMDS ordination space. Two NMDS using
abundance-based (Bray-Curtis) dissimilarity measures were conducted to
map either taxonomic or functional composition, the later based on the 7
leaf, stem and life history traits (without seed mass classes).
Trajectories along time were reported through the Euclidean distance
between the target inventories and the reference inventories in 1989,
\emph{i.e.} 2 years after disturbance. Univariate trajectories of the
leaf, stem and life-history traits were also visualized with the
community weighted means (CWM) \citep{Diaz2007}. Species seed mass
corresponded to 5 classes of increasing mass, seed mass trajectories
were therefore reported as the proportion of each class in the
inventories (Supplementary materials).

The taxonomic diversity was reported through species richness and
evenness, \emph{i.e} the Hill number translation of the Simpson index
\citep{Hill1973}. These indices belong to the set of HCDT or generalized
entropy, respectively corresponding to the 0 and 2 order of diversity
(q), recommended for diversity studies \citep{Marcon2015b}. The
functional diversity was reported using the functional richness and
functional evenness, \emph{i.e} Rao index of quadratic entropy which
combines species abundance distribution and average pairwise
dissimilarity based on species functional traits.

The impacts of initial disturbance were tested with the spearman rank
correlation between the extrema of taxonomic and functional metrics
reached over the 30 years and the initial \%AGB loss. They were besides
analyzed through polynomial regression between (i) taxonomic and
functional richness and evenness and (ii) the initial \%AGB loss at 10,
20 and 30 years after disturbance.

The functional redundancy was measured as the overlap among species in
community functional space \citep{Carmona2016}. The samples of the trait
database were first mapped in a 2-dimensional plan with a PCA analysis.
Then, multivariate kernel density estimator associated with individual
trees returned species traits probability distribution (TDP). Species
TDP weighted by species abundance were eventually summed for each
community. Community functional redundancy was the sum of TDPs overlap,
expressed as the average number of species that could be removed from
without reducing the functional space (Supp. Mat. - Fig. S1 for a more
comprehensive scheme).

@@

\subsection{Communities Composition}\label{communities-composition}

From 1989 (2 years after disturbance) to 2015 (28 years after
disturbance), 828 388 individual trees and 591 botanical species
pertaining to 223 genus and 64 families were recorded.

While both taxonomic and functional composition remained stable in
undisturbed communities (Fig. \ref{fig:NMDSplans}), they followed marked
and consistent trajectories over post-\break disturbance time. In
disturbed communities, these compositional changes corresponded to
shifts towards species with more acquisitive functional strategies, from
communities with high average WSG to high average SLA and chlorophyll
content (see appendix I). For functional composition, this translated
into cyclic compositional changes with an unachieved recovery of the
initial composition (Fig. \ref{fig:NMDSplans}). The maximum
dissimilarity with the initial state was positively correlated to the
disturbance intensity for both taxonomic and functional composition
(\(\rho_{spearman}^{Taxonomic}=0.87\) and
\(\rho_{spearman}^{Functional}=0.90\) respectively). The maximum value
was reached around 26 years after disturbance for taxonomic composition
and 22 years for functional composition.

@@

\caption{Plot trajectories in terms of flora composition (left panels \textbf{(a)} and \textbf{(c)}) and functional composition (right panels \textbf{(b)} and \textbf{(d)}) in a two-dimensional NMDS space. Lower panels (\textbf{(c)} and \textbf{(d)}) represent the Euclidean distance to initial condition along the 30 sampled years. Shaded areas are the credibility intervals.}\label{fig:NMDSplans}
\end{figure*}

Except for leaf chlorophyll content, which continued to increase for
some T3 and T2 plots 30 years after disturbance, all traits and seed
mass proportions followed unimodal trajectories either stabilizing or
returning towards their initial values.

Maximum height at adult stage (\emph{Hmax}), leaf toughness and wood
specific gravity (\emph{WSG}) first decreased and then slightly
increased but remained significantly lower than their initial value
(Fig. \ref{fig:CWM}). On the other side, bark thickness and specific
leaf area (\emph{SLA}) increased and while bark thickness remained
substantially high after 30 years, \emph{SLA} had almost recovered its
initial value. For all traits, the maximum difference to initial value
was correlated to the disturbance intensity
(\(\rho_{spearman}^{Leaf thickness}=0.76\),
\(\rho_{spearman}^{Chlorophyll content}=0.60\),
\(\rho_{spearman}^{Leaf toughness}=-0.53\),
\(\rho_{spearman}^{SLA}=0.93\), \(\rho_{spearman}^{WSG}=-0.75\),
\(\rho_{spearman}^{Bark thickness}=0.71\),
\(\rho_{spearman}^{Hmax}=-0.40\)). The proportions of the three lightest
seed mass classes increased in all disturbed plots, and decreased after
30 years for the lightest class while it stabilized for the two other
(Supp. Mat. - Fig. S2).

@@

\caption{Trajectories of the communities weighted means over 30 years after disturbance of four leaf traits (Leaf thickness, chlorophyll content, toughness, and specific area), two stem traits (wood specific gravity, and bark thickness) and one life history trait (Specific maximum height at adult stage). }\label{fig:CWM}
\end{figure*}

\subsection{Communities richness and
evenness}\label{communities-richness-and-evenness}

For undisturbed plots, taxonomic Richness and Evenness remained stable
over the 30 years monitored. In disturbed communities, after low
disturbance intensity the taxonomic richness increased, reaching a
maximum gain of 14 botanical genera (plot 3 from treatment 2). After
intense disturbance the taxonomic richness followed a more complex
trajectory, decreasing for ten years after disturbance before recovering
to pre-disturbance values. The maximum richness loss or gain after
disturbance was positively correlated to the disturbance intensity
(\(\rho_{spearman}^{Richness}=0.50\)). In all disturbed plots the
taxonomic evenness first increased until a maximum reached after around
20 years. This maximum was positively correlated to the disturbance
intensity (\(\rho_{spearman}^{Evenness}=0.77\)). The evenness then
stabilized except for two T3 plots (plots 8 and 12) for which evenness
kept increasing.

@@

\caption{Trajectories over 30 years of the difference with the 1989 inventory (2 years after disturbance) of community taxonomic \textbf{(a)} richness, \textbf{(b)}, taxonomic evenness, \textbf{(c)} functional richness, and \textbf{(d)} functional evenness. Shaded areas are the credibility intervals }\label{fig:DivTaxo}
\end{figure*}

The plot 7 from treatment 1 displayed constantly outlying functional
richness and evenness and was removed from the graphical representation
for better readability. In undisturbed plots both functional richness
and evenness remained stable along the 30 years. In disturbed plots,
functional richness and evenness trajectories depended on the
disturbance intensity with their maximum positively correlated to \%AGB
loss \(\rho_{spearman}^{Richness}=0.76\) and
\(\rho_{spearman}^{Evenness}=0.60\). Functional richness and evenness
displayed for low disturbance intensity a low but long-lasting increase
up to a maximum reached after 20-25 years, and for high intensity, a
fast but short increase followed after 10 years by a slow decrease
towards the initial values.

The second-degree polynomial regressions between (i) the percentage AGB
loss and (ii) taxonomic and functional richness and evenness after 10,
20 and 30 years best predicted the hump-shaped curve of the disturbance
impact along the disturbance intensity gradient \ref{fig:IDHplot}. The
relationship between the disturbance impact and its intensity was more
markedly hump-shaped for the taxonomic richness than for the taxonomic
evenness. For both functional richness and evenness the relationship was
almost linear. The regression model better predicted the functional
richness and evenness (\(0.55<R^2_{Functional Richness}<0.72\), and
\(0.60<R^2_{Functional Evenness}<0.81\)) than the taxonomic richness and
evenness (\(0.21<R^2_{Taxonomic Richness}<0.4\), and
\(-0.15<R^2_{Taxonomic Evenness}<0.43\) respectively)

>>>>

\Authors{
Ariane MIRABEL\textsuperscript{1*}\\ Bruno Herault\textsuperscript{2}\\ Eric Marcon\textsuperscript{1}
} % Authors
\affiliation{
\textsuperscript{1}UMR EcoFoG, AgroParistech, CNRS, Cirad, INRA, Université des Antilles,
Université de Guyane.\\ \hspace{1em} Campus Agronomique, 97310 Kourou, France.\\\textsuperscript{2}INPHB, Institut National Polytechnique Félix Houphoüet-Boigny\\ \hspace{1em} Yamoussoukro, Ivory Coast.
}
\affiliation{*\textbf{Corresponding author}: ariane.mirabel@ecofog.gf, https://github.com/ArianeMirabel} % Corresponding author

\Keywords{Community Ecology, Community Diversity Determinants, Disturbance Trajectories, Intermediate Disturbance Hypothesis, Mid-term Resilience, Neotropical Forests, Taxonomic and Functional Biodiversity} % Keywords - if you don't want any simply remove all the text between the curly brackets
\newcommand{\keywordname}{Keywords} % Defines the keywords heading name

@@

\Abstract{
Understanding the ecological rules underlying the maintenance of
tropical forests biodiversity, structure, functioning and dynamics is
urgent to anticipate their fate in the global change context. The huge
diversity of tropical forests is often assumed to be regularly reshaped
by natural disturbance yielding a diversity peak at intermediate
intensity. This intermediate disturbance hypothesis (IDH), though,
remains debated and the controversy questions the extent of community
resilience regarding their taxonomic and functional facets. To
disentangle the ecological processes driving community response to
disturbance, we analyzed the tree community trajectories over 30 years
following a disturbance gradient in a Neotropical forest. Specifically,
we examined community functional and taxonomic trajectories with regards
to diversity, composition and redundancy. Functional trajectories were
drawn based on 7 leaf, stem and life-history traits. We highlighted the
cyclic recovery of community taxonomic and functional composition. While
pre-disturbance taxonomic differences were maintained over time,
functional composition trajectories were quite similar among
communities. The IDH did predict community taxonomic diversity response
while functional diversity was enhanced whatever the disturbance
intensity. Although consistent, the recovery of community composition,
diversity and redundancy remained unachieved after 30 years. This
acknowledged the need of decades-long cycles with no disturbance to
ensure a complete recovery, and questioned tropical forest community
resilience after repeated disturbances.
}

@@

The large areas covered with tropical forests worldwide hold crucial
environmental, economic and social values. They provide wood and
multiple non-timber forest products, shelter a diversified fauna,
regulate the local and regional climates, the carbon, water and nutrient
cycles, and ensure cultural and human well-being. The growing demand in
forests products together with current global changes increases the
pressure on remaining natural forests \citep{Morales-Hidalgo2015} and
threatens the maintenance and dynamics in space and time of community
structure, composition and functioning \citep{Anderson-Teixeira2013}.

In tropical forests, ecological communities are regularly re-shaped by
natural disturbance events changing both the abiotic environment,
through the fluxes of light, heat and water \citep{Goulamoussene2017},
and the biotic interactions such as competition among species
\citep{Chesson2000}. One of the cornerstone of tropical forest ecology
is to understand the processes and drivers of ecosystems response to
disturbance \citep{Chazdon2003a}. For now, this has been largely studied
through forest structural parameters such as aboveground biomass, tree
height or stem density \citep{Piponiot2016, Rutishauser2016} that are
rapid and convenient to measure. These structural parameters have been
successfully modeled, giving important insights into the recovery of
ecosystem processes and services \citep{Herault2018}. However the
response of forests diversity and composition remains unclear, albeit it
determines the productivity, stability and functioning of ecosystems
\citep{Tilman2014, Liang2016}. In the short-term, moderate disturbance
may lead to positive impacts on community diversity, an idea formalized
by the intermediate disturbance hypothesis (IDH) which predicts a
maximum species diversity when disturbance intensity is not too high
\citep{Molino2001, Kariuki2006a}.

Validations of the IDH though remain scarce in the long-term and mainly
rely on the analysis of taxonomic richness \citep{Molino2001}. Taxonomic
richness alone, however, gives limited or misleading information on
forests recovery and functioning \citep{Chaudhary2016}. More
ecological-meaningful analysis would couple richness with (i) evenness,
which would reveal the changes in the species abundance distribution and
thus the underlying ecological processes, and (ii) composition, which is
crucial for conservation issues \citep{Lavorel2002, Bellwood2006}.
Furthermore, a functional approach accounting for species biological
attributes would directly link community diversity, composition and
redundancy to ecosystem functioning and to its environmental constraints
\citep{Violle2007b, Baraloto2012a}. In that respect, the functional
trait-based approach that focus on major traits related to species
ecology and mediate species performance in a given environment was
successfully adopted \citep{Diaz2005}. For instance, the functional
approach revealed in tropical rainforests the deterministic processes
entailing, after disturbance, a functional shift from a dominance of
``conservative'' slow-growing species dealing with scarce resources to
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{Reich2014, Herault2011}. This shift is
translated into the trajectories of key functional traits related to
resource acquisition (leaf and stem traits) and life-history traits
(seed mass, maximum size)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}. Eventually,
a complete overview of community response to disturbance would encompass
the changes in functional redundancy, that quantifies the amount of
shared trait values among species \citep{Carmona2016}. The high
functional redundancy of hyper-diverse tropical forests
\citep{Bellwood2006} mitigates the impacts of species removal on
ecosystem functioning and determines the resilience of communities after
disturbance \citep{Elmqvist2003, Diaz2005}.

In this study, we monitored over 30 years the response of 75 ha of
Neotropical forest plots set up on a gradient of disturbance intensity,
from 10 to 60\% of ecosystem above-ground biomass (AGB) loss. We made
use of a large functional traits database encompassing major leaf, stem
and life-history traits in order to draw the taxonomic and functional
trajectories in terms of richness, evenness, composition and redundancy.
Specifically, we (i) elucidated community taxonomic and functional
recovery and the underlying ecological processes, (ii) clarified the
validity of the IDH in the long term for tropical forest and its
translation into different trajectories in time, and (iii) examined
community recovery time.

@@

Paracou station in French Guiana (5\textdegree 18'N and
52\textdegree 53'W) is located in a lowland tropical rain forest in a
tropical wet climate with mean annual temperature of 26\textdegree C,
mean annual precipitation averaging 2980 mm.y\textsuperscript{-1} (30-y
period) and a 3-month dry season (\textless{} 100
mm.month\textsuperscript{-1}) from mid-August to mid-November, and a
one-month dry season in March \citep{Wagner2011}. Elevation ranges
between 5 and 50 m and soils correspond to thin acrisols over a layer of
transformed saprolite with low permeability which produces lateral
drainage during heavy rains.

The experiment is a network of 12 6.25 ha plots that underwent a
disturbance gradient of three logging, thinning and fuelwood cutting
treatments (Table \ref{tab:Tab1}) according to a randomized plot design
with three replicate blocks of four plots. The disturbance corresponds
to an intensity gradient. For treatment 1 (T1) 10 trees of commercial
species (of a diameter at 1.3 m height (DBH) equal or above 50 cm) were
felled per hectare. For treatment 2 (T2) 10 trees/ha of commercial
species (DBH \(\geq\) 50 cm) were felled and 30 trees/ha of non-valuable
species (DBH \(\geq\) 40 cm) were removed by poison girdling. For
treatment 3 (T3) 10 trees/ha of commercial species (DBH \(\geq\) 50 cm)
were felled and 30 trees/ha of non-valuable species (15 with DBH
\(\geq\) 50 cm and 15 with DBH \(\geq\) 40 cm) were removed by poison
girdling \citep{Schmitt1990}. Disturbance intensity was measured as the
percentage of aboveground biomass (\%AGB) lost between the first
inventory in 1984 and five years after disturbance \citep{Piponiot2016}
estimated with the BIOMASS R package \citep{Biomass2018}.

@@

The study site corresponds to a tropical rainforest typical of the
Guiana Shield with a dominance of \emph{Fabaceae},
\emph{Chrysobalanaceae}, \emph{Lecythidaceae} and \emph{Sapotaceae}. In
the 12 experimental plots, all trees above 10 cm DBH have been mapped
and measured annually since 1984. Trees are first identified with a
vernacular name assigned by the forest worker team, and afterward with a
scientific name assigned by botanists during regular botanical
campaigns. In 1984, specific vernacular names were given to 62
commercial or common species whereas more infrequent ones were
identified under general identifiers only distinguishing trees and
palms. From 2003, botanical campaigns have been conducted every 5 to 6
years to identify all trees at the species level but identification
levels still varied among plots and campaigns.

This variability of protocols in time raised methodological issues as
vernacular names usually correspond to different botanical species. This
resulted in significant taxonomic uncertainty that had to be propagated
to composition and diversity metrics. The uncertainty propagation was
done through a Bayesian framework reconstituting complete inventories at
genus level from real incomplete ones on the basis of
vernacular/botanical names association. Vernacular names were replaced
through multinomial trials based on the association probability
\(\big[\alpha_1, \alpha_2,..., \alpha_V\big]\) observed across all
inventories between each vernacular name \emph{v} and all species
\(\big[s_1, s_2,..., s_N\big]\):

@@

Six functional traits representing leaf economics (leaf thickness,
toughness, total chlorophyll content and specific leaf area) and stem
economics (wood specific gravity and bark thickness), and life-history
traits (maximum specific height and seed mass) came from the BRIDGE
project \footnote{http://www.ecofog.gf/Bridge/}. Trait values were
assessed from a selection of individuals located in nine permanent plots
in French Guiana, including two in Paracou, and comprised 294 species
pertaining to 157 genera. Missing trait values (10\%) were filled using
multivariate imputation by chained equation \citep{Mice2011}.
Imputations were restricted within genus or family when samples were too
scarce, in order to account for the phylogenetic signal. Whenever a
species was not in the dataset, it was attributed a set of trait values
randomly sampled among species of the next higher taxonomic level (same
genus or family). As seed mass information was classified into classes,
no data filling process was applied and analyses were restricted to the
414 botanical species recorded.

@@

Because of the variability in the precision of botanical identification
efforts, we were constraint to conduct the taxonomic composition and
diversity analysis at the genus level. Taxonomic and functional
trajectories of community composition were followed in a two-dimensional
NMDS ordination space. Two NMDS using abundance-based (Bray-Curtis)
dissimilarity measures were conducted to map either taxonomic or
functional composition, the latter based on the seven leaf, stem and
life history traits (without seed mass classes). Trajectories along time
were reported through the Euclidean distance between the target
inventories and the reference inventories in 1989, \emph{i.e.} 2 years
after disturbance. Univariate trajectories of the leaf, stem and
life-history traits were also visualized with the community weighted
means (CWM) \citep{Diaz2007}. Species seed mass were given in 5 mass
classes. Seed mass trajectories were therefore reported as the
proportion of each class in the inventories (Supplementary materials).

The taxonomic diversity was reported through species richness and the
Hill number translation of the Simpson index \citep{Hill1973}. The
comparison between richness and Simpson diversity assessed the
variations in taxonomic evenness. Both indices belong to the set of HCDT
or generalized entropy, respectively corresponding to the 0 and 2 order
of diversity (q), recommended for diversity studies \citep{Marcon2015b}.
The functional diversity was reported using the functional richness and
functional evenness, \emph{i.e} the Rao index of quadratic entropy which
combines species abundance distribution and average pairwise
dissimilarity based on species functional traits.

The impacts of initial disturbance were tested with the Spearman rank
correlation between the extremes of taxonomic and functional metrics
reached over the 30 years and the initial \%AGB loss. They were besides
analyzed through polynomial regression between (i) taxonomic and
functional richness and evenness and (ii) the initial \%AGB loss at 10,
20 and 30 years after disturbance.

Functional redundancy was measured as the overlap among species in
community functional space \citep{Carmona2016}. First, the individuals
of the trait database were mapped in the plane of the two first axes
from a PCA analysis. Then, for each species, the traits probability
density (TPD) were computed through multivariate kernel density
estimators. Second, for each community, the functional traits
probability density were computed as the sum of the TDP of the
inventoried species weighted by species abundance. Third, the whole
functional space was divided into a 100 x 100 grid, before counting for
each cell the number of species which TPD is positive. The average count
across cells, minus 1, gives the average number of species in the
community that share the same traits, which is the Community Functional
Redundancy.

@@

\subsection{Community Composition}\label{community-composition}

From 1989 (2 years after disturbance) to 2015 (28 years after
disturbance), 828 388 individual trees and 591 botanical species
pertaining to 223 genera and 64 families were recorded.

While both taxonomic and functional composition remained stable in
undisturbed communities (Fig. \ref{fig:NMDSplans}), they followed marked
and consistent trajectories over post-\break disturbance time. In
disturbed communities, these compositional changes corresponded to
shifts towards species with more acquisitive functional strategies, from
communities with high mean WSG to high mean SLA and chlorophyll content
(see appendix I). For functional composition, this translated into
cyclic compositional changes with an incomplete recovery of the initial
composition (Fig. \ref{fig:NMDSplans}). The maximum dissimilarity with
the initial state was positively correlated with the disturbance
intensity for both taxonomic and functional composition
(\(\rho_{Spearman}^{Taxonomic}=0.87\) and
\(\rho_{Spearman}^{Functional}=0.90\) respectively). The maximum
dissimilarity with the initial was reached for taxonomic composition
between 15 to 25 years, for most of the plots, and around 22 years for
functional composition.

@@

\caption{Plot trajectories in terms of taxonomic composition (\textbf{(a)} and \textbf{(c)}) and functional composition (\textbf{(b)} and \textbf{(d)}) in a two-dimensional NMDS space. Lower panels (\textbf{(c)} and \textbf{(d)}) represent the Euclidean distance to initial condition along the 30 sampled years. Shaded areas are the credibility intervals.}\label{fig:NMDSplans}
\end{figure*}

Community CWM average value of all traits and seed mass proportions
followed unimodal trajectories, either stabilizing or returning towards
their initial values, to the exception of leaf chlorophyll content,
which continued to increase for some T3 and T2 plots 30 years after
disturbance.

Community CWM average value of Maximum height at adult stage
(\emph{Hmax}), leaf toughness and wood specific gravity (\emph{WSG})
first decreased and then slightly increased but remained significantly
lower than their initial value (Fig. \ref{fig:CWM}). On the other side,
bark thickness and specific leaf area (\emph{SLA}) increased and while
bark thickness remained substantially high after 30 years, \emph{SLA}
had almost recovered to its initial value. For all traits, the maximum
difference to initial value was correlated to the disturbance intensity.
Positive correlations were observed for Leaf thickness, chlorophyll
content, SLA and bark thickness
(\(\rho_{Spearman}^{Leaf thickness}=0.76\),
\(\rho_{Spearman}^{Chlorophyll content}=0.60\),
\(\rho_{Spearman}^{SLA}=0.93\),
\(\rho_{Spearman}^{Bark thickness}=0.71\)). Negative correlation was
observed for Leaf toughness, WSG and Hmax
(\(\rho_{Spearman}^{Leaf toughness}=-0.53\),
\(\rho_{Spearman}^{WSG}=-0.75\), \(\rho_{Spearman}^{Hmax}=-0.40\)) The
proportions of the three lightest seed mass classes increased in all
disturbed plots, and decreased after 30 years for the lightest class
while it stabilized for the two other (Supp. Mat. - Fig. S2).

@@

\caption{Trajectories of community weighted means over 30 years after disturbance of four leaf traits (Leaf thickness, chlorophyll content, toughness, and specific area), two stem traits (wood specific gravity, and bark thickness) and one life history trait (Specific maximum height at adult stage). }\label{fig:CWM}
\end{figure*}

\subsection{Community taxonomic and functional
diversity}\label{community-taxonomic-and-functional-diversity}

For undisturbed plots, taxonomic richness and Simpson diversity remained
stable over the 30 years of monitoring. In disturbed communities, after
low disturbance intensity the taxonomic richness increased, reaching a
maximum gain of 14 botanical genera (plot 3 from treatment 2). After
intense disturbance the taxonomic richness followed a more complex
trajectory, decreasing for ten years after disturbance before recovering
to pre-disturbance values. The maximum richness loss or gain after
disturbance was positively correlated with the disturbance intensity
(\(\rho_{Spearman}^{Richness}=0.50\)). In all disturbed plots the
Simpson diversity first increased until a maximum reached after around
20 years. This maximum was positively correlated with the disturbance
intensity (\(\rho_{Spearman}^{Simpson}=0.77\)). The Simpson diversity
then stabilized except for two T3 plots (plots 8 and 12) for which it
kept increasing (Fig. \ref{fig:DivTaxo}).

@@

\caption{Trajectories over 30 years of the difference with the 1989 inventory (2 years after disturbance) of community taxonomic richness \textbf{(a)}, Simpson diversity \textbf{(b)}, functional richness \textbf{(c)}, and Rao diversity \textbf{(d)}. Shaded areas are the credibility intervals }\label{fig:DivTaxo}
\end{figure*}

The plot 7 from treatment 1 displayed constantly outlying functional
richness and Rao diversity and was removed from the graphical
representation for better readability. In undisturbed plots both
functional richness and Rao diversity remained stable along the 30
years. In disturbed plots, both trajectories depended on the disturbance
intensity, with their maximum being positively correlated to \%AGB loss
\(\rho_{Spearman}^{Richness}=0.76\) and \(\rho_{Spearman}^{Rao}=0.60\).
Functional richness and Rao diversity displayed for low disturbance
intensity a low but long-lasting increase up to a maximum reached after
20-25 years, and for high intensity, a fast but short increase followed
after 10 years by a slow decrease towards the initial values.

The second-degree polynomial regressions between (i) the percentage AGB
loss and (ii) taxonomic and functional diversity after 10, 20 and 30
years best predicted the hump-shaped curve of the disturbance impact
along the disturbance intensity gradient \ref{fig:IDHplot}. The
relationship between the disturbance impact and its intensity was more
markedly hump-shaped for the taxonomic richness than for the Simpson
diversity. For both functional richness and Rao diversity the
relationship was almost linear. The regression model better predicted
the functional richness and Rao diversity
(\(0.55<R^2_{Functional Richness}<0.72\), and
\(0.60<R^2_{Functional Rao}<0.81\)) than the taxonomic richness and
evenness (\(0.21<R^2_{Taxonomic Richness}<0.4\), and
\(-0.15<R^2_{Taxonomic Simpson}<0.43\) respectively)
